---
title: "Modeling Global Ecological and Socio-Economic Factors (2015-2023) "
author: "Meron Gedrago, Xiangyu Wang, Adil Gazder, Hongyi Duan"
subtitle: "December "
format: 
  pdf:
    geometry: margin=1in
execute:
  echo: false
  warning: false
---

## Abstract

Our analysis confirmed that larger and wealthier nations are the primary contributors to Carbon Emission, which aligns with their industrial output and economic activities. Encouragingly, some high-income countries are successfully mitigating emissions per capita through progressive policies. For lower-income nations, emissions remain lower but are expected to rise with industrial growth.

Education emerged as a critical factor in economic advancement, with better-educated populations correlating strongly with higher income classifications. This highlights the dual importance of investing in education and sustainable development initiatives.

These findings emphasize the need for global cooperation. Wealthier nations must continue leading emissions reduction efforts, while developing countries require support for sustainable growth. Policymakers should focus on sector-specific interventions, such as clean energy and efficient transportation systems, to reduce environmental impacts globally. Future research should refine these approaches and explore actionable solutions for equitable and sustainable progress.

## Introduction

Carbon emissions, primarily in the form of carbon dioxide (CO₂) released through the burning of fossil fuels, deforestation, and industrial activities, are the primary drivers of global warming. These emissions contribute to the accumulation of greenhouse gases in the atmosphere, trapping heat and causing a rise in global temperatures. The consequences of this warming include more frequent and severe weather events, rising sea levels, and widespread disruption to ecosystems and human livelihoods.

Global warming presents an urgent challenge as its effects are disproportionately felt by low-income nations, which often lack the resources to adapt to climate change. At the same time, high-income countries with historically higher carbon emissions bear significant responsibility for the crisis. This imbalance underscores the necessity for policies that address both environmental sustainability and economic equity. Education plays a crucial role in this effort, equipping individuals with the knowledge and skills to develop innovative solutions, adopt sustainable practices, and influence policy decisions.

Despite growing awareness of the climate crisis, global carbon emissions have continued to rise in recent years. This trajectory threatens progress toward international climate goals, such as those outlined in the Paris Agreement, and underscores the need for deeper exploration of the factors driving emissions and their interconnections with education and income. Addressing these issues requires understanding how education can mitigate carbon emissions and how income inequality influences both education outcomes and environmental impacts.

Building on these concerns, this study aims to investigate the interplay between education, carbon emissions, and income on a global scale. Specifically, it seeks to answer the following questions:

1.  How do National income and population size influence carbon emissions across countries for different compulsory education durations ?

2.  How do life expectancy and labor force affect the the income group of a country, which could indirectly affect carbon emission?

Through an analysis of global datasets encompassing socioeconomic and environmental indicators, this study aims to provide insights into the dynamic relationship between compulsory education, carbon emissions, and socioeconomic characteristics of citizens in these countries.

## Methods

### Data Sourcing and Pre-processing

Our data was a curated dataset from the World Development Indicators (WDI) platform. WDI is the primary World Bank collection of development indicators, compiled from officially recognized international sources. It presents the most current and accurate global development data available, and includes national, regional and global estimates. The database contains 1,400 time series indicators for 217 economies and more than 40 country groups, with data for many indicators going back more than 50 years. We restricted our data to only our variables of interest and associated factors which may affect our research questions.

### Variable Selection

The two main outcome variables we use is the total carbon emissions for a country for a given year and the Income group of a country for a given year. When we write about the variables we will use to estimate the two outcome variables, we have the following restriction on what we refer when we speak on our variables

-   We only restrict the total CO2 emissions to focus only on emissions from direct anthropocentric activities and exclude natural or semi-natural processes.

-   Income Group is defined as the classification of a country into one of the of four categories (Low Income (L), Lower Middle Income (LM), Upper Middle Income (UM) and High Income (H)), based on its Gross Net Income per capita (measured in US\$) for that given year. Based on dynamic thresholds which vary per year, each country is classified into one of the following four categories.

-   National income measures non inflation adjusted and is defined as the difference between Gross National Income and the consumption of fixed capital and natural resources depletion

-   Compulsory Education Duration is the total number of years that children are legally obligated to attend school. For the purpose of this analysis, we have converted this variable into two. If the compulsory education years are below or equal to 10, it is classified as 'Low' and classified as 'High' otherwise.

-   Labor force comprises people ages 15 and older who supply labor for the production of goods and services during a specified period. It includes people who are currently employed and people who are unemployed but seeking work as well as first-time job-seekers. Not everyone who works is included, however. Unpaid workers, family workers, and students are often omitted, and some countries do not count members of the armed forces.

### Model fitting and Evaluation

We chose two different models for our research questions. We used a multiple linear regression to model the carbon emission for countries while we used a multinomial logistic regression model to predict the categories of income group for countries.

-   Multiple linear regression: We conducted an anova test to measure the significance of the interaction term, and tested for linear regression assumptions through the residuals vs fitted and qqplot.

-   Multinomial logistic regression: \[FILL HERE \]

## Results

### Research Question 1: Factors Affecting Carbon Emission

```{r, results='hide'}
# Load necessary libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(MASS)
library(caret)
library(nnet)
#library(summarytools)
library(car)
library(gridExtra)

# Define the URL to the raw file
url <- "https://raw.githubusercontent.com/SamanthaSmiling/stats/main/STATS%20FINAL%20PROJECT%20DATA.xlsx"

# Download the file locally
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, temp_file, mode = "wb")

# Read the Excel file
main_data <- read_excel(temp_file)
main_data$`Labor force ratio` <- main_data$`Labor force, total [SL.TLF.TOTL.IN]`/main_data$`Population, total [SP.POP.TOTL]`

```

```{r, results='hide'}
# Preview the data
glimpse(main_data)
colnames(main_data)
```

```{r, results='hide'}
# Load necessary libraries
# library(dplyr)

# Select relevant columns
data <- main_data[, c("Carbon dioxide (CO2) emissions (total) excluding LULUCF (Mt CO2e) [EN.GHG.CO2.MT.CE.AR5]", 
                     "Compulsory education, duration (years) [SE.COM.DURS]", 
                     "Adjusted net national income (current US$) [NY.ADJ.NNTY.CD]", 
                     "Population, total [SP.POP.TOTL]")]

selected_data <- data %>%
  rename(
    Carbon_Emissions = "Carbon dioxide (CO2) emissions (total) excluding LULUCF (Mt CO2e) [EN.GHG.CO2.MT.CE.AR5]",
    Education_Years = "Compulsory education, duration (years) [SE.COM.DURS]",
    Net_Income = "Adjusted net national income (current US$) [NY.ADJ.NNTY.CD]",
    Population = "Population, total [SP.POP.TOTL]"
  ) %>%
  filter(!is.na(Carbon_Emissions), !is.na(Education_Years), !is.na(Net_Income), !is.na(Population))


selected_data
```

```{r, results='hide'}
#refactoring education years into two categories 
selected_data <- selected_data %>% mutate(Education_new_fac = case_when(
  Education_Years <=10 ~ "Low",
  Education_Years >10 ~ "High",
    TRUE ~ "Other"
  ))

```

```{r, results='hide'}
# Summary statistics
summary(selected_data)
summary(selected_data$Education_new_fac)

# Convert Education_Years to a factor
#selected_data$Education_Years <- as.factor(selected_data$Education_Years)

# Scatter plot matrix
#pairs(selected_data, main = "Scatter Plot Matrix")

```

```         
```

The data as shown below in 'Table 1' shows big differences between countries in carbon emissions, income, education, and population. A few countries produce a lot of carbon emissions, raising the global average, while most emit much less. Income is also uneven, with some very rich countries pulling the average far above what most others earn. Education levels are more even, with most countries requiring about 10 years of school, but improving the quality of education may help poorer nations. Bigger, richer countries tend to have larger populations and produce more carbon, but population size alone doesn’t explain the differences. These patterns show that solving problems like climate change and poverty needs different plans for different countries—rich, high-emission countries need to lead the way in cutting emissions and sharing resources to help others grow in a fair and sustainable way.

Table 1: Characteristics of country carbon emission and demography

|              | Carbon Emission | Compulsory Education | National Income (billion USD) | Population (million) |
|---------------|---------------|---------------|---------------|---------------|
| Median       | 39.39           | 10                   | 82.93                         | 17.34                |
| 1st Quartile | 7.14            | 9                    | 13.22                         | 4.39                 |
| 3rd Quartile | 447.70          | 11                   | 944.3                         | 126.7                |
| Mean         | 1584.21         | 9.78                 | 2651                          | 391.3                |

When we further explore the relationship of the four variables, we can see a strong positive association exists between Population and Carbon Emission, indicating that countries with larger populations tend to emit more carbon, likely due to increased industrial and energy demands. Similarly, National Income shows a positive trend with Carbon , suggesting that wealthier countries might contribute more to emissions. For countries with lower Compulsory Education, the relationship between National Income and Carbon Emission is stronger when compared to countries with higher Compulsory Education. Overall, all three variables look to be related to Carbon Emission from the scatterplot.

```{r}
# Visualize relationships
library(ggplot2)

ggplot(selected_data, aes(x = Net_Income, y = Carbon_Emissions, 
                          color = Education_new_fac)) +
  geom_point(aes(size=Population)) +  # Size mapped to points here
  geom_smooth(method = "lm", se = FALSE) +  # Smooth line doesn't need size
  labs(
    title = "Carbon Emissions vs. Net Income",
    color = "Education Level",
    size = "Population size"
  ) +
  theme_minimal() +
  theme(legend.position = "right")



```

```{r, results='hide'}
# Build the model
carbon_model_interac <- lm(Carbon_Emissions ~ Education_new_fac * Net_Income + Population, data = selected_data)

# Summary of the model
summary(carbon_model_interac)

```

Table 2: Multiple Linear Regression Model Summary

| Variable                                | Estimate  | SE        | t-value  | p-value |
|-----------------------------------------|-----------|-----------|----------|---------|
| Compulsory Education                    | -110.7    | 61.44     | -1.504   | 0.133   |
| National Income                         | 2.040e-10 | 1.034e-11 | 19.728   | \<0.001 |
| Population                              | 2.471e-6  | 4.666e-08 | 52.966   | \<0.001 |
| Compulsory Education \* National Income | 8.714e-11 | 1.168e-11 | 7.463    | \<0.001 |

The regression model shows a strong overall fit, indicating that 93.52% of the variation in carbon emissions is explained by Compulsory Education, National Income and Population.The regression model shows how factors like education, income, and population affect carbon emissions. It tells us that net income and population size are very important in predicting carbon emissions—higher income and more people lead to more carbon emissions. Education level alone doesn't seem to have a big effect, but when combined with income, it changes the way income affects emissions. Specifically, countries with lower education seem to have a stronger link between income and carbon emissions. This means that while income and population are the biggest drivers, education can still influence how these factors interact with emissions.

```{}
```

```{r}
# Residual diagnostics
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 layout
#plot(carbon_model_factor)
plot(carbon_model_interac)

```

```{r, results='hide'}

# Compare the models using ANOVA with the model without the interaction 
# Build the model - without interaction 
carbon_model <- lm(Carbon_Emissions ~ Education_new_fac + Net_Income + Population, data = selected_data)

# Summary of the model
summary(carbon_model)

#anova test 
anova(carbon_model, carbon_model_interac)

```

```{r, results='hide'}
# Residual diagnostics
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 layout
#plot(carbon_model_factor)
plot(carbon_model)
```

The factor model highlights that the relationship between education years and carbon emissions is nuanced, with certain education levels (like 8 and 8.5 years) being particularly impactful. This could indicate critical thresholds where education systems or societal behaviors influence emissions. For policy-making, focusing on specific education levels rather than increasing education years linearly might yield better outcomes for carbon reduction strategies.While both models perform well, the factor model provides deeper insights and better predictive power, making it more suitable for nuanced analyses and interventions.

### Research Question 2: Factors affecting Income Group

```{r, warning=FALSE}


eda_data <- main_data[, c("Income Group", "Time", "Population, total [SP.POP.TOTL]",
                        "Life expectancy at birth, total (years) [SP.DYN.LE00.IN]", 
                        "Adjusted net national income (current US$) [NY.ADJ.NNTY.CD]")]

eda_data <- eda_data[!(is.na(eda_data$`Income Group`) | eda_data$`Income Group` == "") &
             !(is.na(eda_data$`Life expectancy at birth, total (years) [SP.DYN.LE00.IN]`) | 
               eda_data$`Life expectancy at birth, total (years) [SP.DYN.LE00.IN]` == "") &
             !(is.na(eda_data$`Adjusted net national income (current US$) [NY.ADJ.NNTY.CD]`) | 
               eda_data$`Adjusted net national income (current US$) [NY.ADJ.NNTY.CD]` == ""), ]


p1 <-ggplot(eda_data, aes(y = `Life expectancy at birth, total (years) [SP.DYN.LE00.IN]`, 
                      x = log(`Adjusted net national income (current US$) [NY.ADJ.NNTY.CD]`), 
                      color = `Income Group`)) +
  geom_point(size = 3, alpha = 0.8) + # Scatter points with size and transparency
  geom_smooth(method = "lm", se = FALSE, aes(group = `Income Group`), linewidth = 0.8) + # Regression lines without confidence intervals
  labs(title = "Life Expectancy vs. Log Net National Income",
       x = "Log Adjusted Net National income",
       y = "Life Expectancy",
       color = "Income Group") + # Labels for axes and legend
  theme_minimal() # A clean theme for the plot

p1

```

```{r, warning=FALSE}

eda_data <- main_data[, c("Income Group", "Time", "Population, total [SP.POP.TOTL]",
                        "Life expectancy at birth, total (years) [SP.DYN.LE00.IN]", 
                        "Labor force ratio")]

eda_data <- eda_data[!(is.na(eda_data$`Income Group`) | eda_data$`Income Group` == "") &
             !(is.na(eda_data$`Life expectancy at birth, total (years) [SP.DYN.LE00.IN]`) | 
               eda_data$`Life expectancy at birth, total (years) [SP.DYN.LE00.IN]` == "") &
             !(is.na(eda_data$`Labor force ratio`) | 
               eda_data$`Labor force ratio` == ""), ]


p2 <-ggplot(eda_data, aes(x = `Life expectancy at birth, total (years) [SP.DYN.LE00.IN]`, 
                      y = `Labor force ratio`, 
                      color = `Income Group`)) +
  geom_point(size = 3, alpha = 0.8) + # Scatter points with size and transparency
  geom_smooth(method = "lm", se = FALSE, aes(group = `Income Group`), linewidth = 0.8) + # Regression lines without confidence intervals
  labs(title = "Life Expectancy vs. Labor Force Ratio",
       x = "Life Expectancy at Birth (Years)",
       y = "Labor force ratio",
       color = "Income Group") + # Labels for axes and legend
  theme_minimal() # A clean theme for the plot

```

```{r, warning=FALSE}


eda_data <- main_data[, c("Income Group", "Time", "Population, total [SP.POP.TOTL]",
                        "Life expectancy at birth, total (years) [SP.DYN.LE00.IN]", 
                        "Labor force ratio")]

eda_data <- eda_data[!(is.na(eda_data$`Income Group`) | eda_data$`Income Group` == "") &
             !(is.na(eda_data$`Life expectancy at birth, total (years) [SP.DYN.LE00.IN]`) | 
               eda_data$`Life expectancy at birth, total (years) [SP.DYN.LE00.IN]` == "") &
             !(is.na(eda_data$`Population, total [SP.POP.TOTL]`) | 
               eda_data$`Population, total [SP.POP.TOTL]` == ""), ]


p3 <-ggplot(eda_data, aes(x = `Life expectancy at birth, total (years) [SP.DYN.LE00.IN]`, 
                      y = log(`Population, total [SP.POP.TOTL]`), 
                      color = `Income Group`)) +
  geom_point(size = 3, alpha = 0.8) + # Scatter points with size and transparency
  geom_smooth(method = "lm", se = FALSE, aes(group = `Income Group`), linewidth = 0.8) + # Regression lines without confidence intervals
  labs(title = "Life Expectancy vs. Log Total Population",
       x = "Life Expectancy at Birth (Years)",
       y = "Log (Total Population)",
       color = "Income Group") + # Labels for axes and legend
  theme_minimal() # A clean theme for the plot

```

```{r, warning=FALSE, results='hide'}
# Choose the data we need
data <- main_data[, c("Income Group", 
                     "Time", 
                     "Life expectancy at birth, total (years) [SP.DYN.LE00.IN]", 
                     "Labor force, total [SL.TLF.TOTL.IN]", 
                     "Population, total [SP.POP.TOTL]")]

colnames(data) <- c("Income Group",
                    "Year",
                    "Life Expectancy (Years)",
                    "Total Labor Force",
                    "Total Population")

data$`Income Group` <- factor(data$`Income Group`, 
                              levels = c("L", "LM", "UM", "H"), 
                              ordered = TRUE)

data$`Labor Force Ratio` <- data$`Total Labor Force`/data$`Total Population`
data <- na.omit(data)
                
str(data) 
```

```{r, warning=FALSE}
# The box-plot of Life Expectancy (Years) and Labor Force Ratio for different Income Group
p3 <- ggplot(data, aes(x = `Income Group`, y = `Life Expectancy (Years)`)) +
  geom_boxplot(fill = "green") +
  labs(title = "Boxplot of Life Expectancy by Income Group", x = "Income Group", y = "Life Expectancy (Years)")
p4 <- ggplot(data, aes(x = `Income Group`, y = `Labor Force Ratio`)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of Labor Force Ratio by Income Group", x = "Income Group", y = "Labor Force Ratio")

grid.arrange(p3, p4, ncol = 2)
```

```{r, warning=FALSE}
# Joint Distribution Plots of Life Expectancy and Labor Force Ratio
p5 <- ggplot(data, aes(x = `Life Expectancy (Years)`, y = `Labor Force Ratio`)) +
  geom_point() +
  stat_smooth(method = "lm") +
  labs(title = "Joint Distribution with Linear Fit")
```

```{r, results='hide'}
# Normalize the data
cols <- c("Life Expectancy (Years)", "Total Labor Force")
data[cols] <- scale(data[cols])

# Add an intersection term for Case 3
data$`Intersection Term` <- data$`Life Expectancy (Years)` * data$`Labor Force Ratio`
```

```{r, results='hide'}
# Random divide the data in 80% to 20% Ratio
set.seed(123)
indices <- 1:nrow(data)
train_indices <- sample(indices, size = floor(0.8 * length(indices)))
test_indices <- setdiff(indices, train_indices)
train_data <- data[train_indices, ]
test_data <- data[test_indices, ]
```

```{r, warning=FALSE, results='hide'}
model5 <- polr(`Income Group` ~ `Life Expectancy (Years)` + `Labor Force Ratio`, 
               data = train_data, Hess = TRUE)
summary(model5)
```

```{r, warning=FALSE, results='hide'}
coefficients5 <- coef(model5)
Odds_Ratio5 <- exp(coefficients5)
pred5 <- predict(model5, test_data)
conf_matrix5 <- confusionMatrix(pred5, test_data$`Income Group`)
conf_matrix5
```

```{r, results='hide'}
conf_matrix_percent_row5 <- round(prop.table(conf_matrix5$table, 1) * 100, 2)
conf_matrix_percent_row5
```

## Conclusion

The project aimed to uncover the drivers of global CO2 emissions and explore the relationship between socioeconomic indicators and a country’s income classification. Using the World Development Indicators (WDI) dataset, we analyzed variables such as population size, income levels, education years, and urbanization rates to address two key questions. This study offers valuable insights into the global dynamics of CO2 emissions and their link to socioeconomic indicators. By analyzing comprehensive data from the World Bank, we investigated the impact of variables like population size, income, and education on both emissions and income classification.

Through a systematic process of data sourcing, preprocessing, variable selection, and rigorous statistical analysis—including linear and non-linear regression models—we identified meaningful patterns.⁠ According to the statistical analysis ⁠, we found that income and population size are the most influential predictors of CO2 emissions, with wealthier and larger countries contributing disproportionately. Education, while less directly linked to emissions, plays a crucial role in how National Income impacts Carbon Emission. These results underscore the interplay between economic activity and environmental impact, highlighting the need for tailored strategies to address emissions across different income groups. ⁠

Future research should integrate industry-specific data, such as emissions from energy and transportation sectors, and leverage advanced modeling techniques to deepen understanding. This could guide policy interventions aimed at balancing economic development with environmental sustainability.

## References
